Draft Letter

### 2024-03-13 rawfiles/Draft_letter.qmd
first big update after asking for feedback. Thanks everyone across many plaforms
lunchtime Thursday- an added paragraph and a little caption rewriting

Things David knows he still needs to do:

@@

-   standardised citation format.

@@

@@

-   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of check names all[^2].
-   That Gibson's method, producing positive excess mortality, is more effective than Karlinsky & Kobak in determining mortality as it adjusts for population change due to immigration changes.

@@

The Our World in Data graph, and the Karlinsky and Kobak method that underlies it is an approximation (if it is a better or worse approximation to Gibson's method will be discussed later). While recognising the Our World In Data graph is an approximation, there are still several practically useful implicit features about it.

@@

While the existence of centuries of mortality measurement (not mentioned in Gibson's article) does not directly address if New Zealand medical professionals are aware that best practice mortality calculations recognise that people of different ages die at different rates when using the publicly available Our World in Data graphs in media pieces, it does establish that there is an established methodology, not mentioned in Gibson's article, underpinning claims of New Zealand's negative excess, and that method has been used for centuries, world-wide, across multiple disciplines. We can check if New Zealand medical professionals are aware of this by asking them: *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ’s are!)"*(Ashley Bloomfield) [^6]

@@

Some variation in calculation results can be caused by the choice of standard population, due to the variation of young and old within a population. However, pragmatically, any population that can be argued to be analytically useful should produce similar results. Using single year ages groups by sex to 94 then 95+, and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -5% for 2020 to 2022. Using a standard reference population of 2021 when the borders were restricted gives a cumulative excess of -6%. Using a pre-COVID-19 standard population of 2019 gives a cumulative excess of -5%. Using the standard population of 1961, used by Stats NZ to maintain long-term continuity, cumulative excess is -5%.

@@

![baseline](md_figures/figure2.png){fig-alt="Overlong baseline error" fig-align="center"}

Based on the influenza season patterns in recent pre-pandemic years, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline[^7]. New Zealand has seen a very similar history in this period to Australia, with the added limitation that incorporating 2011, the year of the Christchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is a poor influence point to use in a baseline as it is not a re-occurring annual event.

[^7]: Australian Bureau of Statistics. https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology

@@

![Even if all years are before COVID-19, the post-baseline years are above trend](md_figures/figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}

>>>>

Things David knows he still needs to do (that don't alter overall content):

@@

-   standard citation format.

@@

-   Captions under graphs.

@@

-   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of Karlinsky & Kobak[^2].
-   That Gibson's method, producing positive excess mortality, is more effective than Karlinsky & Kobak in determining mortality as it adjusts for population change due to immigration changes.

@@

The Our World in Data graph, and the Karlinsky and Kobak method that underlies it, is an approximation (if it is a better or worse approximation to Gibson's method will be discussed later). While recognising the Our World In Data graph is an approximation, there are still several practically useful implicit features about it:

@@

While the existence of centuries of mortality measurement (not mentioned in Gibson's article) does not directly address if New Zealand medical professionals are aware that best practice mortality calculations recognise that people of different ages die at different rates when using the publicly available Our World in Data graphs in media pieces, it does establish that there is an established methodology, not mentioned in Gibson's article, underpinning claims of New Zealand's negative excess, and that method has been used for centuries, world-wide, across multiple disciplines. We can check if New Zealand medical professionals are aware of this by asking them: *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ’s are!)"*(Ashley Bloomfield) [^6].

@@

Some variation in calculation results can be caused by the choice of standard population, due to the variation of young and old within a population. However, pragmatically, any population that can be argued to be analytically useful should produce similar results. Using single year ages groups by sex to 94 then 95+[^7], and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -5% for 2020 to 2022. Using a standard reference population of 2021 when the borders were restricted gives a cumulative excess of -6%. Using a pre-COVID-19 standard population of 2019 gives a cumulative excess of -5%. Using the standard population of 1961, used by Stats NZ to maintain long-term continuity, cumulative excess is -5%.

[^7]: NZ annual deaths are available from <https://infoshare.stats.govt.nz> in Population : Deaths : Deaths by age and sex (Annual-Dec). NZ mean annual population is available from Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec) : Mean year ended.

@@

![Figure 2: Structural error introduced by overly long baselines. Longterm (age) standardised death rates from https://infoshare.stats.govt.nz in Population : Death Rates - DMM : Standardised death rates (Maori and total population) (Annual-Dec) : Total Population](md_figures/figure2.png){fig-alt="Overlong baseline error" fig-align="center"}

Based on the influenza season patterns in recent pre-pandemic years, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline[^8]. New Zealand has seen a very similar history in this period to Australia, with the added limitation that using 2011, the year of the Christchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is an extremely poor choice to include in baselines as an earthquake hitting the second largest city and people living in broken housing through the winter is not a reoccuring annual event.

[^8]: Australian Bureau of Statistics. https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology

@@

![Even if all years are before COVID-19, the post-baseline years are above trend](md_figures/figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"} Of more surprise is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the accepted method of comparing death rates while ignoring the effects of age. While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same answers, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of the regression. But, compared to the differences with methods acknowledging age, such problems are minor.
